---
title: "Online Retail Exploratory Data Analysis"
author: "Charles Aniji"
date: "2025-04-01"
output:
  html_document: default
  pdf_document: default
  word_document: default
editor_options: 
  chunk_output_type: inline
---

## 1. Setup

We’ll use **data.table** for speed on ~500 K rows, plus **dplyr**, **lubridate**, and **arules** for analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(data.table)
library(dplyr)
library(lubridate)
library(arules)
library(tidyverse)
```  

## 2. Load Data

Read the cleaned CSV into a `data.table`.

```{r}
online_retail <- fread("C:/Users/ADM1N/Desktop/UK_Retail_Analytics/online_retail_cleaned.csv", header = TRUE, fill = Inf)
```  

## 3. Inspect Data

### 3.1 Dimensions

```{r}
dim(online_retail)
```  

### 3.2 Preview

```{r}
head(online_retail, 10)
```  

### 3.3 Structure

```{r}
str(online_retail)
```  


## 4. Prepare Date Fields

Convert `InvoiceYearMonth` string to a Date object for grouping.


```{r echo=FALSE}
online_retail$InvoiceYearMonth <- ym(online_retail$InvoiceYearMonth)
```  


## 5. Summary Statistics

Compute mean, min, max, and standard deviation for **Quantity**, **UnitPrice**, and **TotalSpent**.


```{r summary}
summary_stats <- online_retail %>% 
  summarise(
    Mean_Quantity    = mean(Quantity,    na.rm = TRUE),
    Min_Quantity     = min(Quantity,     na.rm = TRUE),
    Max_Quantity     = max(Quantity,     na.rm = TRUE),
    SD_Quantity      = sd(Quantity,      na.rm = TRUE),
    Mean_UnitPrice   = mean(UnitPrice,   na.rm = TRUE),
    Min_UnitPrice    = min(UnitPrice,    na.rm = TRUE),
    Max_UnitPrice    = max(UnitPrice,    na.rm = TRUE),
    SD_UnitPrice     = sd(UnitPrice,     na.rm = TRUE),
    Mean_TotalSpend  = mean(TotalSpent,  na.rm = TRUE),
    Min_TotalSpend   = min(TotalSpent,   na.rm = TRUE),
    Max_TotalSpend   = max(TotalSpent,   na.rm = TRUE),
    SD_TotalSpend    = sd(TotalSpent,    na.rm = TRUE)
  ) %>% 
  round(2)

print(summary_stats)
```  


## 6. Top Products

### 6.1 By Sales Volume


```{r}
top_products_volume <- online_retail %>%
  group_by(Description) %>%
  summarise(Total_Units_Sold = sum(Quantity, na.rm = TRUE)) %>%
  arrange(desc(Total_Units_Sold)) %>%
  slice_max(n = 10, Total_Units_Sold)

print(top_products_volume)
```  


### 6.2 By Revenue


```{r}
top_products_revenue <- online_retail %>%
  group_by(Description) %>%
  summarise(Total_Revenue = sum(TotalSpent, na.rm = TRUE)) %>%
  arrange(desc(Total_Revenue)) %>%
  slice_max(n = 10, Total_Revenue)

print(top_products_revenue)
```  


## 7. Repeat Purchase Rate

Calculate the percentage of customers with more than one distinct order.


```{r}
repeat_customers <- online_retail %>%
  group_by(CustomerID) %>%
  summarise(OrderCount = n_distinct(InvoiceNo)) %>%
  summarise(
    RepeatCustomers = sum(OrderCount > 1),
    TotalCustomers  = n(),
    RepeatRate      = round(100 * RepeatCustomers / TotalCustomers, 1)
  )

print(
  glue::glue(
    "Repeat purchase rate: {repeat_customers$RepeatRate}%"
  )
)
```  


## 8. Monthly Sales Trend

### 8.1 Aggregate Revenue by Month


```{r}
monthly_sales <- online_retail %>%
  group_by(InvoiceYearMonth) %>%
  summarise(Total_Revenue = sum(TotalSpent, na.rm = TRUE)) %>%
  arrange(InvoiceYearMonth)

print(monthly_sales)
```  


### 8.2 Identify Peak Month


```{r}
peak_month <- monthly_sales %>%
  filter(Total_Revenue == max(Total_Revenue)) %>%
  pull(InvoiceYearMonth)

print(glue::glue("Highest-revenue month: {peak_month}"))
```  


### 8.3 Top 3 Products in Peak Month


```{r}
top_peak_products <- online_retail %>%
  filter(InvoiceYearMonth == peak_month) %>%
  group_by(Description) %>%
  summarise(Monthly_Revenue = sum(TotalSpent, na.rm = TRUE)) %>%
  arrange(desc(Monthly_Revenue)) %>%
  slice_max(n = 3, Monthly_Revenue)

print(top_peak_products)
```  


## 9. Market Basket Analysis

### 9.1 Prepare Transactions


```{r}
transactions_list <- split(online_retail$Description, online_retail$InvoiceNo)
transactions      <- as(transactions_list, "transactions")
```  


### 9.2 Run Apriori & Filter


```{r}
basket_rules <- apriori(
  transactions,
  parameter = list(supp = 0.014, conf = 0.77)
)

filtered_rules <- subset(basket_rules, lift > 1.5)
filtered_rules <- sort(filtered_rules, by = c("lift", "confidence"), decreasing = TRUE)

inspect(head(filtered_rules, 5))
```  


### 9.3 Convert Rules to a Data Frame


```{r}
rules_df <- as(filtered_rules, "data.frame")
head(rules_df)
```


### 9.4 Split Into LHS and RHS


```{r}
rules_df <- rules_df %>%
  separate(
    col    = rules,
    into   = c("lhs", "rhs"),
    sep    = "=>",
    remove = FALSE
  ) %>%
  mutate(
    lhs = str_remove_all(lhs, "[\\{\\}]") %>% trimws(),
    rhs = str_remove_all(rhs, "[\\{\\}]") %>% trimws()
  )

head(rules_df)
```


## 10. RFM Segmentation

### 10.1 Compute RFM Metrics


```{r}
reference_date <- max(online_retail$InvoiceDate)

rfm_data <- online_retail %>%
  group_by(CustomerID) %>%
  summarise(
    Recency   = as.numeric(reference_date - max(InvoiceDate)),
    Frequency = n_distinct(InvoiceNo),
    Monetary  = sum(TotalSpent, na.rm = TRUE)
  )
```  


### 10.2 Score & Segment


```{r}
rfm_scored <- rfm_data %>%
  mutate(
    R_Score   = ntile(-Recency, 5),
    F_Score   = ntile(Frequency, 5),
    M_Score   = ntile(Monetary, 5),
    RFM_Group = case_when(
      R_Score >= 4 & F_Score >= 4 & M_Score >= 4 ~ "Champions",
      R_Score >= 3 & F_Score >= 3 & M_Score >= 3 ~ "Loyal Customers",
      R_Score >= 4 & F_Score <= 2 & M_Score <= 2 ~ "Recent Buyers",
      R_Score <= 2 & F_Score >= 4 & M_Score >= 4 ~ "At Risk Customers",
      R_Score <= 2 & F_Score <= 2 & M_Score <= 2 ~ "Lost Customers",
      TRUE                                      ~ "Others"
    )
  )

rfm_summary <- rfm_scored %>%
  group_by(RFM_Group) %>%
  summarise(
    Avg_Recency   = mean(Recency),
    Avg_Frequency = mean(Frequency),
    Avg_Monetary  = mean(Monetary),
    Customer_Count = n()
  )

print(rfm_summary)
```  


## 11. Export for Tableau
### Save the RFM summary and rules for dashboarding

```{r}
fwrite(rfm_summary, "C:/Users/ADM1N/Desktop/UK_Retail_Analytics/rfm_summary.csv")
fwrite(rules_df, "C:/Users/ADM1N/Desktop/UK_Retail_Analytics/filtered_rules.csv")
fwrite(online_retail, "C:/Users/ADM1N/Desktop/UK_Retail_Analytics/online_retail_viz_ready.csv")
```  
